Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-05-01 manuscript/manuscript.tex
ms updates

Looking a neutral simulations results, we found that spuriously significant
$G(t)$ trajectories (e.g. different from zero under the case of neutrality) can
be created by conditioning on sojourn time. Our intuition, which will need to
be confirmed by future work and theory, is that by conditioning on a quick
sojourn time to fixation (or loss) from an initial intermediate frequency,
leads to trajectories that, purely by chance, have many upward (or downward)
allele frequency changes, which leads to a positive sum of pairwise temporal
covariances (the numerator of $G(t)$). Similarly, conditioning on long sojourn
times implies that frequency changes alternate back and forth quite a bit, such
the sum of pairwise temporal covariances could be slightly negative.

When all pairwise covariances are calculated and
summed to calculate the numerator of $G(t)$, this leads to this subset of
neutral trajectories to have spuriously significant and positive $G(t)$
trajectories. We briefly confirmed that conditioning on fixation leads to a
spurious positive numerator of $G(t)$. Looking at one neutral simulation, we
conditioned on alleles fixing during the 105 generations of which temporal
covariances are calculated. Per locus, we calculated the sum of all pairwise
products between allele frequency changes (analogous to the numerator of
$G(t)$). Whether we condition on fixation, or a large absolute allele frequency
change between the initial generation and the last generation, we find that
quickly fixing sites and sites with large allele frequency shifts have a
significant positive sum of covariances as expected, despite being neutral
trajectories. Thus, even under neutrality, one must mind how fixations are
handled.

Since our simulation allele
frequencies are population, rather than sample frequencies, when an allele is
fixed or lost is known with certainty.

  \caption{$G(t)$ trajectories (left) and temporal covariances (right) from
    truncation selection simulations for different numbers of individuals
    selected (line color). Dashed lines indicate $G(t)$ trajectories and
    temporal covariances calculated \emph{including} fixed sites, while the
    solid lines exclude fixed sites. All values are averaged over 30 replicate
    simulations; the lines in the right figure are loess smoothed, while points
    are averages. The solid lines of the temporal covariances have been
    excluded in the left figure for clarity, but are similar except they do not
    become negative.}

First, we found that under strong truncation selection (i.e. the number of
selected individuals is small), our measure $G(t)$ can become negative. This
possibility was discussed in \textcite{Buffalo2019-io} as a result of
fluctuating selection; here, however, the buildup of negative covariances
driving this

Since our simulations were focused on understanding how genetic architecture
and mode of selection impact temporal covariance and $G(t)$ (with a limited
number of neutral sites, due to computational limitations), population
frequencies were used.

If frequencies are measured without sampling noise, including fixed or lost
sites in the calculation of temporal covariance and $G(t)$ has the effect of
shrinking the covariance estimates towards zero,

how fixations (i.e. when a site is polymorphic but is fixed or
lost during the course of its trajectory) are handled. Through simulations and
comparing the results of analyzing our empirical datasets both ways, we have
found trade-offs to both approaches, but overall find that not removing fixed
sites is a more conservative, safer approach, and is what we use throughout our
paper.

Temporal covariances (and related statistics like convergence correlations,
$G(t)$, etc.) are calculated on allele frequency changes, $\Delta p_t$; when an
allele is fixed or lost, the allele frequency changes are zero thereafter
$\Delta p_t = 0$. When temporal covariance is calculated by averaging over the
loci in a region or across the genome, fixed sites zero out the terms in the
sum of the right hand side of temporal covariance, $\cov(\Delta p_t, \Delta
p_s) = \E(\Delta p_t \Delta p_s) - \E(\Delta p_t) \E(\Delta p_s)$ and can
shrink the second terms towards zero. From simulations using population
frequencies (e.g. no sampling noise), we have found that including neutral
sites into the temporal covariance calculation that are lost or fix will tend
to shrink the covariance towards zero, and in some cases create spurious
negative temporal covariances. We observed this when conducting the background
selection simulations of Supplementary Materials Section XXX, and have included
an alternative version of this figure that includes $G(t)$ trajectories
calculated by including fixed neutral sites (Supplementary Figure
\ref{suppfig:supp-bgs-g-fix}). We find, as expected, $G(t)$ trajectories
calculated including fixed sites are shrunk towards zero, and in some cases,
can become negative. In contrast, the $G(t)$ trajectories ignoring fixations
(dashed lines in Supplementary Figure \ref{suppfig:supp-bgs-g-fix}) are higher.

XXX truncation selection, exponential.

While simulation results using population frequencies show that including fixed
sites has the downside of shrinking the temporal covariance towards zero, and
in some cases can create negative covariances, we have found excluding fixed
sites on sample frequency data has a severe drawback: low frequency alleles
will occasionally not be observed at a sample timepoint (leading to a frequency
of zero of the minor allele), and excluding these observations (instead of
treating it as a trajectory that has a 0 frequency timepoint) biases estimates.
We observed this bias created by ignoring fixed or loss sites by comparing our
$N_e$ estimates for the \textcite{Barghi2019-qy} study (estimated from the
total variance in allele frequency change) with the estimates from the original
paper. We found that ignoring fixed or lost sites (by treating them as missing
data in calculating the covariances) led to many low-frequency alleles not
contributing to the variance, leading this statistic to be calculated on more
intermediate frequency alleles and thus biasing the $N_e$ estimate downwards.
Additionally, we tried only dropping fixed or loss sites from the temporal
covariance calculations that were at the end or the beginning of a trajectory
(e.g., as if the site was created by a new mutation or fixed); while this
ameliorated some of this bias, it still did not lead to $N_e$ estimates
congruent with the original studies. Overall, we found by trying all these
approaches that not removing fixed or lost sites was the best way to deal with
sample allele frequencies that could be missing from some timepoints.

Still, we wanted to ensure that our empirical re-analyses were robust to our
choice of how we handled fixed sites. Given that under some simulation
conditions, we find that spurious negative covariances can occur when fixations
are included, we wanted to ensure that our result that the temporal covariances
of \textcite{Barghi2019-qy} that become negative at later timepoints are not an
artifact of the way fixations are handled. First, we regenerated Figure 1 A and
B, except with temporal covariances and $G(t)$ calculated excluding fixed or
lost sites. We find the same qualitative results regardless of whether we
include or exclude fixed/lost sites; importantly, we still see a significant
signal that the temporal covariances at later timepoints become negative. As
expected the temporal covariances are lessened by excluding fixed sites.

XXX

@@

  \caption{The effect of including fixed/lost sites in the calculation of the
  temporal covariances and $G(t)$ trajectories of the \textcite{Barghi2019-qy}
data. Dashed lines are those including fixed/lost sites (i.e. the original
Figure 1), and solid lines are excluding fixed/lost sites.}

@@

>>>>

Looking a neutral simulations results, we found that spuriously significant
$G(t)$ trajectories (e.g. different from zero under the case of neutrality) can
be created by conditioning on sojourn time. Our intuition, which will need to
be confirmed by future work and theory, is that by conditioning on a quick
sojourn time to fixation (or loss) from an initial intermediate frequency,
leads to trajectories that, purely by chance, have many upward (or downward)
allele frequency changes, which leads to a positive sum of pairwise temporal
covariances (the numerator of $G(t)$). Similarly, conditioning on long sojourn
times implies that frequency changes alternate back and forth quite a bit, such
the sum of pairwise temporal covariances could be slightly negative. However,
the overall affect could also be affected by the average allele frequency
change across loci and other factors.

To further complicate the matter, our simulation allele frequencies are
population, rather than sample frequencies, when an allele is fixed or lost is
known with certainty. With empirical data calculated on sample allele
frequencies, low frequency minor alleles many not be sampled at some
timepoints, and excluding these observations (instead of treating it as a
trajectory that has a 0 frequency timepoint) biases estimates.  We observed
this bias created by ignoring fixed or loss sites by comparing our $N_e$
estimates for the \textcite{Barghi2019-qy} study (estimated from the total
variance in allele frequency change) with the estimates from the original
paper. We found that ignoring fixed or lost sites (by treating them as missing
data in calculating the covariances) led to many low-frequency alleles not
contributing to the variance, leading this statistic to be calculated on more
intermediate frequency alleles and thus biasing the $N_e$ estimate downwards.
Additionally, we tried only dropping fixed or loss sites from the temporal
covariance calculations that were at the end or the beginning of a trajectory
(e.g., as if the site was created by a new mutation or fixed); while this
ameliorated some of this bias, it still did not lead to $N_e$ estimates
congruent with the original studies. Overall, we found by trying all these
approaches that not removing fixed or lost sites was the best way to deal with
sample allele frequencies that could be missing from some timepoints.

@@

  \caption{The effect of excluding fixed/lost sites in the calculation of the
  temporal covariances and $G(t)$ trajectories of the \textcite{Barghi2019-qy}
data. Dashed lines are those including fixed/lost sites (i.e. the original
Figure 1), and solid lines are excluding fixed/lost sites.}

@@

To ensure that our findings were robust to handling sites with a frequency of
zero or one differently, we regenerated Figures \ref{fig:figure-1} and Figure
\ref{fig:figure-3} but excluded frequencies of zero or one. Specifically, we
wanted to ensure that our finding that later timepoints had negative
covariances at later timepoints was not spuriously caused by the way fixed or
lost sites are handled. We see no qualitative difference (Supplementary
Material Figures \ref{suppfig:supp-fig-1-nofix} and
\ref{suppfig:supp-fig-3-nofix}) that emerges when sites with frequency zero or
one are excluded.

  \caption{$G(t)$ trajectories (A) and temporal covariances (B) from
    truncation selection simulations for different numbers of individuals
    selected (line color). Dashed lines indicate $G(t)$ trajectories and
    temporal covariances calculated \emph{including} fixed sites, while the
    solid lines exclude fixed sites. All values are averaged over 30 replicate
    simulations; the lines in the right figure are loess smoothed, while points
    are averages. The solid lines of the temporal covariances have been
    excluded in the left figure for clarity, but are similar except they do not
    become negative.}

    \label{suppfig:supp-trunc}
\end{figure}

Through simulations, we noticed the strongest effect fixation  or loss of sites
has on temporal covariance and $G(t)$ is under truncation selection. Here,
since only a (potentially small) fraction of individuals contribute to the next
generation, sites can fix over very short timescales. Furthermore, the small
number of effective breeders contributing to the next generation shrinks $N_e$
considerably, which increases $\var(p_t - p_0$, the denominator of $G(t)$. We
see the effect of handling fixed/lost sites differently, and the faster rate of
drift in Supplementary Materials Figure \ref{suppfig:supp-trunc} (A), where
weaker truncation selection actually has higher levels of $G(t)$. Looking just
at temporal covariances, we find that stronger truncation selection (e.g.  a
smaller tail of individuals selected) does lead to greater temporal
covariances.
